---
title: "Time in Hospital for Diabetes Patients"
author: "Leslie Vazquez Moreno"
date: "2024-03-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(caret)
```

```{r}
# diabetes data
diabetes = read.csv("~/Desktop/UCR Winter 2024/STAT208/project/data/diabetes.csv")

diabetes[diabetes == "?"] <- NA
diabetes[diabetes == "Unknown/Invalid"] <- NA
```

# Question/Problem

Which variables are important in predicting the length of stay at a hospital for a diabetic patient? 

How well does the prediction classifier model perform?

# EDA

```{r}
diabetes = diabetes[4:ncol(diabetes)]

par(mfrow = c(1,2))

# time_in_hospital
barplot(table(diabetes[,"time_in_hospital"], useNA = "ifany"), 
        main = "Barplot Time in the Hospital",
        xlab = "Days in the Hospital",
        ylab = "Frequency",
        cex.names = 0.4)
# insulin
barplot(table(diabetes[,"insulin"], useNA = "ifany"), 
        main = "Barplot Insulin",
        xlab = "Insulin Given/Dosage Change",
        ylab = "Frequency")
par(mfrow = c(1,1))

# dev.print(pdf, 'common_var.pdf')
```

```{r}
par(mfrow = c(2,2))
# weight
na_count = sum(is.na(diabetes[,"weight"]))
non_na = length(diabetes[,"weight"]) - na_count
barplot(c(non_na, na_count), main = "Barplot NA in Weight",
        names.arg = c("non-NA", "NA"),
        xlab = "Weight", ylab = "Frequency")
# payer_code
na_count = sum(is.na(diabetes[,"payer_code"]))
non_na = length(diabetes[,"payer_code"]) - na_count
barplot(c(non_na, na_count), main = "Barplot NA in Payer Code",
        names.arg = c("non-NA", "NA"),
        xlab = "Payer Code", ylab = "Frequency")
# medical_specialty
na_count = sum(is.na(diabetes[,"medical_specialty"]))
non_na = length(diabetes[,"medical_specialty"]) - na_count
barplot(c(non_na, na_count), main = "Barplot NA in Medical Specialty",
        names.arg = c("non-NA", "NA"),
        xlab = "Medical Specialty", ylab = "Frequency")
par(mfrow = c(1,1))

# dev.print(pdf, 'na_var.pdf')
```

```{r}
par(mfrow = c(1,2))
# examide
barplot(table(diabetes[,"examide"], useNA = "ifany"), 
        main = "Barplot Examide",
        ylab = "Frequency",
        xlab = "Examide Given/Dosage Change")
# citoglipton
barplot(table(diabetes[,"citoglipton"], useNA = "ifany"), 
        main = "Barplot Citoglipton",
        ylab = "Frequency",
        xlab = "Citoglipton Given/Dosage Change")
par(mfrow = c(1,1))

# dev.print(pdf, 'same_var.pdf')
```

```{r}
diabetes2 = diabetes[, -c(4, 5, 6, 7, 38, 39)]
diabetes2 = na.omit(diabetes2)
# apply(diabetes2, 2, unique)
diabetes2 = diabetes2[,-c(38, 27, 32, 37, 24)]
# apply(diabetes2, 2, unique)

for(i in c(13:15)){
  diabetes2[,i] = chartr("V", "1", diabetes2[,i])
  diabetes2[,i] = chartr("E", "1", diabetes2[,i])
}
diabetes2[,19:34] <- ifelse(diabetes2[,19:34] == "No", 0, 1)

diabetes3 = diabetes2 %>%
  mutate(race = ifelse(race == "Caucasian", 1, 0),
         gender = ifelse(gender == "Male", 1, 0),
         age = as.factor(age), # as.numeric? binary 60 year threshold?
         time_in_hospital = ifelse(time_in_hospital <= 3, 1, 0),
         payer_code = ifelse(payer_code == "SP", 1, 0),
         medical_specialty = ifelse(medical_specialty == "Pediatrics-Endocrinology" | medical_specialty == "InternalMedicine" | medical_specialty == "Endocrinology" | medical_specialty == "Endocrinology-Metabolism", 1, 0),
         diag_1 = ifelse(floor(as.numeric(diag_1)) == 250, 1,0),
         diag_2 = ifelse(floor(as.numeric(diag_2)) == 250, 1,0),
         diag_3 = ifelse(floor(as.numeric(diag_3)) == 250, 1,0),
         max_glu_serum = ifelse(max_glu_serum == "None", 0,
                                ifelse(max_glu_serum == "Norm", 1, 2)),
         A1Cresult = ifelse(A1Cresult == "None", 0,
                                ifelse(A1Cresult == "Norm", 1, 2)),
         change = ifelse(change == "Ch", 1, 0),
         diabetesMed = ifelse(diabetesMed == "Yes", 1, 0),
         readmitted = ifelse(readmitted == "NO", 0,
                             ifelse(readmitted == "<30", 1, 2)))
```

```{r}
# time_in_hospital, drug variables
par(mfrow = c(1,2))
barplot(table(diabetes3[,"time_in_hospital"], useNA = "ifany"), 
        main = "Barplot Time in the Hospital",
        ylab = "Frequency",
        xlab = "Less Than 4 Days in Hospital")
barplot(table(diabetes3[,"insulin"], useNA = "ifany"), 
        main = "Barplot Insulin",
        xlab = "Insulin Given",
        ylab = "Frequency")
par(mfrow = c(1,1))

# dev.print(pdf, 'new_var.pdf')
```

Chi-Square Independence Test with time_in_hospital
```{r}
# independence test
p_value = c()
cat_var = c(1:3, 5:ncol(diabetes3))
for(i in cat_var){
  p_value[which(cat_var == i)] = as.numeric(chisq.test(diabetes3[, i], diabetes3[,"time_in_hospital"])[3])
}
chi_y = data.frame(variable = colnames(diabetes3[, c(1:3, 5:ncol(diabetes3))]), 
                   p_value,
                   independence = ifelse(p_value < 0.05, "No", "Yes"))
knitr::kable(chi_y[chi_y$independence == "Yes",])
# outputs only variables with an independent interpretation
```

Correlation Heat Map
```{r, fig.width = 5, fig.height = 5}
library(ggcorrplot)
model.matrix(time_in_hospital~., data=diabetes3) %>% 
  cor(use="pairwise.complete.obs") %>% 
  ggcorrplot(show.diag=FALSE, type="lower", lab=TRUE, lab_size=0.8) + 
  theme(axis.text.x=element_text(size=4),
        axis.text.y=element_text(size=4)) +
  ggtitle("Correlation Matrix Between Variables")

# dev.print(pdf, 'corr.pdf')
```

# Model

Logistic Regression: Significant Coefficients
```{r}
# Logistic regression
library(glmnet)
library(boot)
log_reg = glm(time_in_hospital ~ ., data = diabetes3, family = "binomial")
summary(log_reg)[5] # aic
# summary(log_reg)$coefficients[,1]
sig_index = as.numeric(which(summary(log_reg)$coefficients[,4] < 0.05))
knitr::kable(data.frame(coefficients = summary(log_reg)$coefficients[sig_index[1:11],1], p_value = summary(log_reg)$coefficients[sig_index[1:11],4]))
```

```{r}
knitr::kable(data.frame(coefficients = summary(log_reg)$coefficients[sig_index[12:21],1], p_value = summary(log_reg)$coefficients[sig_index[12:21],4]))
```

CV estimate of prediction error
```{r}
cv.glm(diabetes3, log_reg, K = 10)$delta[1]
```

Splitting into Train and Test

Logistic Regression
```{r}
# Logistic regression
# splitting data 20% test, 80% train
set.seed(1)
test_index = sample(1:nrow(diabetes3), nrow(diabetes3)/5)
test = diabetes3[test_index, ]
train = diabetes3[-test_index, ]

library(glmnet)
library(boot)
log_reg = glm(time_in_hospital ~ ., data = train, family = "binomial")
summary(log_reg)[5] # aic
# summary(log_reg)$coefficients[,1]
# sig_index = as.numeric(which(summary(log_reg)$coefficients[,4] < 0.05))
# knitr::kable(data.frame(coefficients = summary(log_reg)$coefficients[sig_index[1:11],1], p_value = summary(log_reg)$coefficients[sig_index[1:11],4]))
test_resp = predict(log_reg, newdata = test, type = "response")
test_pred <- ifelse(test_resp > 0.5, 1, 0)
mean(test_pred != test$time_in_hospital)
```

Logistic Regression subset (using AIC)
```{r}
# # backwards
log_reg_aic = step(log_reg)
# summary(log_reg_aic)$call
# summary(log_reg_aic)$coefficients
```

```{r}
log_reg_aic = glm(time_in_hospital ~ race + gender + age + payer_code + medical_specialty + num_lab_procedures + num_procedures + num_medications + number_outpatient + number_emergency + number_inpatient + diag_1 + diag_2 + diag_3 + number_diagnoses + max_glu_serum + A1Cresult + metformin + repaglinide + glimepiride + glyburide + pioglitazone + insulin + diabetesMed + readmitted, data = train, family = "binomial")
summary(log_reg_aic)[5] 
summary(log_reg_aic)$coefficients

test_resp = predict(log_reg_aic, newdata = test, type = "response")
test_pred <- ifelse(test_resp > 0.5, 1, 0)
mean(test_pred != test$time_in_hospital)
```

```{r}
varImp(log_reg_aic) %>%
  arrange(desc(Overall))

V = varImp(log_reg_aic)
ggplot2::ggplot(V, aes(x=reorder(rownames(V),Overall), y=Overall)) +
geom_col()+
xlab('Variable')+
ylab('Overall Importance')+
  ggtitle("Logistic Regression Variable Importance") + 
theme_bw() +
  theme(axis.text.y=element_text(size=4)) + 
  coord_flip()


# dev.print(pdf, 'lr_importance.pdf')
```

Random Forest Full
```{r}
# Random forest
library(randomForest)
rf1 = randomForest(as.factor(time_in_hospital) ~ ., data = train)
rf1[5]
```

```{r}
pred_test1 <- predict(rf1, newdata = test, type= "class")
mean(pred_test1 != test$time_in_hospital)

confusionMatrix(table(pred_test1,test$time_in_hospital))
```

```{r}
V_r = varImp(rf1)
ggplot2::ggplot(V_r, aes(x=reorder(rownames(V_r),Overall), y=Overall)) +
geom_col()+
xlab('Variable')+
ylab('Overall Importance')+
  ggtitle("Random Forest Variable Importance") + 
theme_bw() +
  theme(axis.text.y=element_text(size=4)) + 
  coord_flip()

# dev.print(pdf, 'rf_importance.pdf')
```

```{r}
library(RColorBrewer)
heatmap(rf1[[5]][1:2, 1:2], Colv = NA, Rowv = NA, scale = "none",
main = "Random Forest Confusion Matrix",
col = colorRampPalette(brewer.pal(8, "Blues"))(25),
margins = c(0.7, 0.7))
legend("topright", c("Small", "Large"),
fill=colorRampPalette(brewer.pal(8, "Blues"))(3),
     cex = 0.7)
# dev.print(pdf, 'rf_conf_mat.pdf')
```

Random Forest Subset

```{r}
# Random forest
# only variables scoring less than 13 are removed
rf2 = randomForest(as.factor(time_in_hospital) ~ race + gender+ age+payer_code+medical_specialty + num_lab_procedures + num_procedures + num_medications + number_outpatient + number_emergency + number_inpatient + diag_1 + diag_2 + diag_3 + number_diagnoses + max_glu_serum + A1Cresult + metformin + repaglinide + nateglinide + glimepiride + glipizide + glyburide + pioglitazone + rosiglitazone + acarbose + insulin + glyburide.metformin + change + diabetesMed + readmitted, data = train)
rf2[5]
```

```{r}
pred_test2 <- predict(rf2, newdata = test, type= "class")
mean(pred_test2 != test$time_in_hospital)

confusionMatrix(table(pred_test2,test$time_in_hospital))

varImp(rf2)  %>%
  arrange(desc(Overall))
```

# Results

```{r}
# 5 most important variables
# logistic regression aic chosen model
varImp(log_reg_aic)  %>%
  arrange(desc(Overall)) %>%
  top_n(5)

# random forest full model
varImp(rf1)  %>%
  arrange(desc(Overall)) %>%
  top_n(5)
```


```{r}
# 5 least important variables
# logistic regression aic chosen model
varImp(log_reg_aic)  %>%
  arrange(desc(Overall)) %>%
  top_n(-5)

# random forest full model
varImp(rf1)  %>%
  arrange(desc(Overall)) %>%
  top_n(-5)
```

```{r}
# logistic regression aic chosen model
mean(test_pred != test$time_in_hospital)

# random forest full model
mean(pred_test1 != test$time_in_hospital)
```


# Dataset

<https://www.kaggle.com/datasets/jimschacko/10-years-diabetes-dataset/data>

Variable Information

<https://www.hindawi.com/journals/bmri/2014/781670/tab1/>

<https://www.hindawi.com/journals/bmri/2014/781670/tab2/>


